# `readxl` {#readxl}

## Summary

**Check this, may need to be a block quote**: The **readxl** package makes it easy to import tabular data from Excel spreadsheets (.xls or .xlsx files) and includes several options for cleaning data during import. **readxl** has no external dependencies and functions on any operating system, making it an OS- and user-friendly package that simplifies getting your data from Excel into R. 

## Objectives

- Use `readr::read_csv()` to read in a comma separated value (CSV) file
- Use `readxl::read_excel()` to read in an Excel worksheet from a workbook
- Replace a specific string/value in a spreadsheet with with `NA` 
- Skip *n* rows when importing an Excel worksheet
- Use `readxl::read_excel()` to read in parts of a worksheet (by cell range)
- Specify column names when importing Excel data
- Read in multiple Excel worksheets
- Something about readxl workflows (e.g. %>% `write_csv()` to keep a copy of the raw data in a project)?? 

## Resources

- https://readxl.tidyverse.org/


## Lesson

### Lesson prep: get data files into your working directory

In Session 1, we introduced how and why R Projects are great for reproducibility, because our self-contained working directory will be the **first** place R looks for files. 

You downloaded four files for this workshop: 

- fish_counts_curated.csv
- invert_counts_curated.xlsx
- kelp_counts_curated.xlsx
- substrate_cover_curated.xlsx

Copy and paste those files into the 'RforExcelUsers' folder on your computer. Notice that now these files are in your working directory when you go back to that Project in RStudio (check the 'Files' tab). That means they're going to be in the first place R will look when you ask it to find a file to read in. 

### Create a new .R script, attach the `tidyverse` and `readxl` packages  

In your RforExcelUsers project in RStudio, open a new .R script and add a useful header to the top of it. For example, at a minimum: 

```{r}

# --------------------------------
# A descriptive title
# Summary of what this script is for 
# Your name
# Contact information
# --------------------------------

# Other things you might include: required packages or datasets, relevant links (e.g. to raw data source, GitHub repo, etc.), citations and sources.
```

In this lesson, we'll read in a CSV file with the `readr::read_csv()` function, so we need to have the `readr` package attached. Since it's part of the `tidyverse`, we'll go ahead and attach the `tidyverse` package below our script header using `library(package_name)`. It's a good idea to attach all necessary packages near the top of a script, so we'll also attach the `readxl` packages here. 

```{r, message = FALSE}
# Attach the tidyverse
library(tidyverse)

# Attach the readxl package
library(readxl)
```

Now, all of the packages and functions within the `tidyverse` and `readxl`, including `readr::read_csv()` and `readxl::read_excel()`, are available for use. 

### Use `readr::read_csv()` to read in data from a CSV file

There are many types of files containing data that you might want to work with in R. A common one is a comma separated value (CSV) file, which contains values with each column entry separated by a comma delimiter. CSVs can be opened, viewed, and worked with in Excel just like an .xls or .xlsx file - but let's learn how to get data directly from a CSV into R where we can work with it more reproducibly. 

The CSV we'll read in here is called "fish_counts_curated.csv", and contains observations for "the abundance and size of fish species as part of SBCLTER's kelp forest monitoring program to track long-term patterns in species abundance and diversity" from the [Santa Barbara Channel Long Term Ecological Research](http://sbc.lternet.edu/) program. 

**Source:** Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Fish abundance. Environmental Data Initiative. https://doi.org/10.6073/pasta/dbd1d5f0b225d903371ce89b09ee7379. Dataset accessed 9/26/2019.

Read in the "fish_counts_curated.csv" file `read_csv("file_name.csv")`, and store it in R as an object called *fish_counts*:

```{r, include = FALSE}

# Teaching version
fish_counts <- read_csv("curation/fish_counts_curated.csv")

```

```{r, eval = FALSE, results = 'hide'}

fish_counts <- read_csv("fish_counts_curated.csv")

```

Notice that the name of the stored object (here, *fish_counts*) will show up in our Environment tab in RStudio. 

Click on the object in the Environment, and R will automatically run the `View()` function for you to pull up your data in a separate viewing tab. Now we can look at it in the spreadsheet format we're used to. 

Let's make a simple plot of some fish counts with `ggplot2`. 

```{r}
ggplot(fish_counts, aes(x = year, y = tot_count, group = common_name)) +
  geom_col(aes(fill = common_name)) +
  facet_wrap(~site)
```

Now that we have our fish counts data ready to work with in R, let's get the substrate cover and kelp data (both .xlsx files). In the following sections, we'll learn that we can use `readxl::read_excel()` to read in Excel files directly.

### Use `readxl::read_excel()` to read in a single Excel worksheet 

First, take a look at *substrate_cover_curated.xlsx* in Excel, which contains a single worksheet with substrate type and percent cover observations at different sampling locations in the Santa Barbara Channel. 

A few things to notice:

- The file contains a single worksheet
- There are multiple rows containing text information up top
- Where observations were not recorded, there exists '-9999'

Let's go ahead and read in the data. If the file is in our working directory, we can read in a single worksheet .xlsx file using `readxl::read_excel("file_name.xlsx")`. *Note: readxl::read_excel() works for both .xlsx and .xls types*. 

Like this: 
```{r, include = FALSE}

substrate_cover <- read_excel("curation/substrate_cover_curated.xlsx")

```

```{r, eval = FALSE, results = 'hide'}
substrate_cover <- read_excel("substrate_cover_curated.xlsx")
```

**Tada? Not quite.** 

Click on the object name (*substrate_cover*) in the Environment to view the data in a new tab. A few things aren't ideal:

- The top row of text has automatically become the (messy) column headers
- There are multiple descriptive rows before we actually get to the data
- There are -9999s that we want R to understand `NA` instead

We can deal with those issues by adding arguments within `read_excel()`. Include argument `skip = n` to skip the first 'n' rows when importing data, and `na = "this"` to replace "this" with `NA` when importing:

```{r, include = FALSE}
substrate_cover <- read_excel("curation/substrate_cover_curated.xlsx", skip = 4, na = "-9999")
```

```{r, eval = FALSE}
substrate_cover <- read_excel("curation/substrate_cover_curated.xlsx, skip = 4, na = "-9999")
```

Check out *substrate_cover*, and see that the first row *after* the 4 skipped are the column names, and all -9999s have been updated to `NA`. Hooray!

So far we've read in a single CSV file using `readr::read_csv()`, and an Excel file containing a single worksheet with `readxl::read_excel()`. Now let's read in data from an Excel file with multiple worksheets. 

### Use `readxl::read_excel()` to read in selected worksheets 

Now, we'll read in the kelp fronds data from file *kelp_counts_curated.xlsx*. If you open the Excel workbook, you'll see that it contains multiple worksheets with giant kelp observations in the Santa Barbara Channel during July 2016, 2017, and 2018, with data collected at each *site* in a separate worksheet.

To read in a single Excel worksheet from a workbook we'll again use `readxl::read_excel("file_name.xlsx")`, but we'll need to let R know which worksheet to get. 

Let's read in the kelp data just like we did above, as an object called *kelp_counts*.

```{r, include = FALSE}

# Teaching chunk (won't show up when knitted)
kelp_counts <- readxl::read_excel("curation/kelp_counts_curated.xlsx")

```

```{r, eval = FALSE, results = 'hide'}
kelp_counts <- readxl::read_excel("kelp_counts_curated.xlsx")
```

You might be thinking, "Hooray, I got all of my Excel workbook data!" But remember to always look at your data - you will see that actually only the first worksheet was read in. The default in `readxl::read_excel()` is to read in the **first worksheet** in a multi-sheet Excel workbook. 

If we want to read in data from a worksheet other than the first one in an Excel workbook, we can specify the correct worksheet by name or position by adding the `sheet` argument. 

Let's read in data from the worksheet named *golb* (Goleta Beach) in the *kelp_counts_curated.xlsx* workbook: 
```{r, include = FALSE}
kelp_golb <- readxl::read_excel("curation/kelp_counts_curated.xlsx", sheet = "golb")
```

```{r, eval = FALSE, results = 'hide'}
kelp_golb <- readxl::read_excel("kelp_counts_curated.xlsx", sheet = "golb")
```

Note that you can also specify a worksheet by position: since *golb* is the 6^th^ worksheet in the workbook, we could also use the following: 
```{r, include = FALSE}
kelp_golb <- readxl::read_excel("curation/kelp_counts_curated.xlsx", sheet = 6)
```

```{r, eval = FALSE}
kelp_golb <- readxl::read_excel("kelp_counts_curated.xlsx", sheet = 6)
```


### Use readxl::read_xls() to read in pieces of an Excel worksheet


### Use purrr::map() and readxl::read_xls() to read in multiple worksheets simultaneously


## Fun facts (quirky things) - making a note of these wherever possible for interest (little “Did you know?” sections)
## Interludes (deep thoughts/openscapes)
## Our Turn Your Turn 1
## Our Turn Your Turn 2
## Efficiency Tips
