---
title: "Kelp counts data curation"
author: "Allison Horst"
date: "9/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(writexl)
```


Curating the 'annual_kelp_all_years.csv' LTER data into an Excel worksheet with multiple tabs (stored as kelp_counts.xlsx). 

**Info:** https://portal.edirepository.org/nis/mapbrowse?scope=knb-lter-sbc&identifier=18&revision=newest

**Citation:** Reed D. 2018. SBC LTER: Reef: Kelp Forest Community Dynamics: Abundance and size of Giant Kelp (Macrocystis Pyrifera), ongoing since 2000. Environmental Data Initiative. https://doi.org/10.6073/pasta/dea56e02127161194626f97c8b6118e8. Dataset accessed 9/05/2019.

1. Get the raw data: 
```{r}

kelp_data <- read_csv("annual_kelp_all_years.csv")

```

2. Split it up by SITE (will put data for each of 11 sites in a separate Excel worksheet, within the same workbook)
```{r}

kelp_split <- kelp_data %>% 
  group_by(SITE) %>% 
  group_split()
```

```{r}

# Just get the names & arrange (in order of list appearance)
site_names <- kelp_data %>% 
  select(SITE) %>% 
  arrange(SITE)

# Set the names of the kelp_split list items
names(kelp_split) <- tolower(unique(site_names$SITE))

# Write to an Excel file: 
write_xlsx(kelp_split, "kelp_counts.xlsx")
```

